---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
options(width = 120)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# RITCH - an R interface to the ITCH Protocol

<!-- badges: start -->
[![Travis build status](https://travis-ci.org/DavZim/RITCH.svg?branch=master)](https://travis-ci.org/DavZim/RITCH)
<!-- badges: end -->

## The What?!
This R package allows you to read files that use the ITCH protocol (version 5.0) of NASDAQ and parse it into a data.table. 

The ITCH protocol allows NASDAQ to distribute financial information to market participants. The financial information includes orders, trades, order modifications, trading status, traded stocks, and more.

## The Why?!

During my research I had to parse some ITCH files and couldn't find any R libraries. While parsing the files in pure R is certainly possible, this library uses `C++` and `Rcpp` to speed up the parsing process. 

A typical file containing a single trading day consists of something like 30-50 million messages (BX-exchange) up to 230 million messages (NASDAQ), thus speed makes a crucial difference. As the data is streamed from the file, the execution time mainly depends on the reading/writing speed of the hard-drive. 

As a general benchmark, it takes my machine about 20 seconds to count the messages of a plain-file (unzipped, 55 million order) and about 10 seconds longer for a `.gz`-file.

Enough of the talk, how can I use the package?


## The How?!

Currently the code only lives on GitHub but eventually it may find its way to CRAN. Until then, you have to use `remotes` or `devtools` to install `RITCH`:

### Installation
You can install the development version from [GitHub](https://github.com/DavZim/RITCH) with:

``` r
# install.packages("remotes")
remotes::install_github("DavZim/RITCH")
```

### Example Usage

As a first step, we want to count how often each message is found in a given file.

### Counting Messages

RITCH is able to read gzipped (`*.XXX_ITCH_50.gz`) and unzipped files (`*.XXX_ITCH_50`), for speed reasons, we gunzipped the file by hand as we will use it multiple times throughout this example.

As I don't own any rights to the data, the data is not included in this package, but can be downloaded from NASDAQs FTP server here: ftp://emi.nasdaq.com/ITCH/

```{r}
library(RITCH)

file <- "20191230.BX_ITCH_50"

msg_count <- count_messages(file, add_meta_data = TRUE)
msg_count
```

As you can see, there are a lot of different message types. Currently this package parses only messages from the following groups
- "Add Order Messages" (type 'A' and 'F'), 
- "Modify Order Messages" (type 'E', 'C', 'X', 'D', and 'U'), 
- and "Trade Messages" (type 'P', 'Q', and 'B'). 

To get an overview of this information, you can also use the `get_meta_data()` function. 
```{r}
get_meta_data()
```

You can extract the different message-types by using the functions `get_orders()`, `get_modifications()`, and `get_trades()`, respectively. The doc-number refers to the section in the official documentation (see link below).

If you are annoyed by the feedback the function gives you (`[Counting] ... [Converting]...`), you can always turn the feedback off with the `quiet = TRUE` option (this applies to all functions of this package).

If you need more message type compatibility, you are more than welcome to post an issue or open a pull request.

### Retrieve Orders
```{r}
orders  <- get_orders(file)
orders
```

If you want to load only a specified number of messages (this applies to all `get_*` functions), you can always specify a start and end message number.

For example, if you want to get only the first 10 orders, you can use the following code.

```{r}
orders_small  <- get_orders(file, 1, 10)
orders_small
```

### Retrieve Trades
```{r}
trades <- get_trades(file)
trades
```

### Retrieve Order Modifications
```{r}
mods <- get_modifications(file)
mods
```

To speed up the `get_*` functions, we can use the message-count information from earlier. For example the following code yields the same results as above, but saves time.

```{r}

orders <- get_orders(file, msg_count)
trades <- get_trades(file, msg_count)
mods   <- get_modifications(file, msg_count)
# # alternatively, provide the start and end number of messages:
# orders <- get_orders(file, 1, count_orders(msg_count))
# trades <- get_trades(file, 1, count_trades(msg_count))
# mods   <- get_modifications(file, 1, count_modifications(msg_count))
```


### Create a Plot with Trades and Orders of the largest ETFs

As a last step, a quick visualisation of the data

```{r ETF_plot, cache=TRUE}
library(ggplot2)

# load the data
orders <- get_orders(file, msg_count)
trades <- get_trades(file, msg_count)

# data munging
tickers <- c("SPY", "IWO")
dt_orders <- orders[stock %in% tickers]
dt_trades <- trades[stock %in% tickers]

# for each ticker, use only orders that are within 1% of the range of traded prices
ranges <- dt_trades[, .(min_price = min(price), max_price = max(price)), by = stock]
# filter the orders
dt_orders <- dt_orders[ranges, on = "stock"][price >= 0.99 * min_price & price <= 1.01 * max_price]
# replace the buy-factor with something more useful
dt_orders[, buy := ifelse(buy, "Bid", "Ask")]
dt_orders[, stock := factor(stock, levels = tickers)]

# data visualization
ggplot() +
  # add the orders to the plot
  geom_point(data = dt_orders, 
             aes(x = as.POSIXct(datetime), y = price, color = buy), size = 0.5, alpha = 0.2) +
  # add the trades as a black line to the plot
  geom_step(data = dt_trades, 
            aes(x = as.POSIXct(datetime), y = price)) +
  # add a facet for each ETF
  facet_grid(stock~., scales = "free_y") +
  # some Aesthetics
  theme_light() +
  labs(title = "Orders and Trades of the largest ETFs",
       subtitle = "Date: 2019-12-30 | Exchange: BX", 
       caption = "Source: NASDAQ ITCH",
       x = "Time", y = "Price", 
       color = "Side") +
  scale_y_continuous(labels = scales::dollar) +
  scale_color_brewer(palette = "Set1")
```


## Some considerations

All functions that take a file-name work for both `.gz`-files (i.e., `YYYYMMDD.XXX_ITCH_50.gz`) or or plain-files (`YYYYMMDD.XXX_ITCH_50`). The compressed files will be uncompressed into a temp-file and deleted after the function finishes, if you want to use multiple commands on one file, it might be faster to use `R.utils::gunzip(..., remove = FALSE)` in R or `gunzip -k YYYYMMDD.XXX_ITCH_50.gz` in the terminal and call the functions on the plain, uncompressed file.

Parsing another example file (`20170130.BX_ITCH_50.gz`, 714MB gzipped and 1.6GB unzipped) has a peak RAM-consumption of around 7GB on my machine.

If the file is too large for your RAM, you can also process everything in batches by providing a start and end message count. I.e., to only parse the first 1,000 orders, you can use `get_orders(file, 1, 1000)`.

To speed the parsing up, you can also specify the number of messages. I.e., count all messages once using `count_messages()`, and then provide the number of trades/orders/order-modifications to `end_msg_count` in each subsequent function call. This saves the parser one trip over the file. 

You can also just provide the message counts directly, e.g.:
```{r}
msg_count <- count_messages(file)
orders <- get_orders(file, msg_count)
```
versus providing the actual start and end position of messages:
```{r}
msg_count <- count_messages(file)
orders <- get_orders(file, 1, count_orders(msg_count))
```

## Additional Sources aka. Data

While this package does not contain any real financial data using the ITCH format, NASDAQ provides some sample datasets on its FTP-server, which you can find here: [ftp://emi.nasdaq.com/ITCH](ftp://emi.nasdaq.com/ITCH)

If you want to find out more about the protocol, have a look at the official protocol specification, which you can find here: https://www.nasdaqtrader.com/content/technicalsupport/specifications/dataproducts/NQTVITCHspecification.pdf

If you find this package useful or have any other kind of feedback, I'd be happy if you let me know. Otherwise, if you need more functionality for additional message types, please feel free to create an issue or a pull request. 

## Open Issues

To move the package towards CRAN, I want to include a smaller data file containing fake or simulated data, this needs to be converted to the ITCH format. This would allow the example code to run, but also to properly use unit tests in the package.

Currently the `C++` code parses the data into data.frame, which is then converted into a data.table. If possible, the data.table should be directly populated. But I haven't found the time to look into a Rcpp data.table API. 
